hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2024-12-09 inst/manuscript/paper.Rmd
fix merge conflicts in manuscript

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`. Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for `hatchR`. The first is a R package distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on `hatchR`'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

`hatchR` is meant to primarily be a tool for predicting phenology. In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualization from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice. Finally, we provide brief summaries of general applications of `hatchR` below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables.

@@

The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though `model_table` includes more complex models developed by @beacham1990), as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has includes parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

Fundamentally, `hatchR` assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade). Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required. We expect your data to look something like this:

@@

`hatchR` assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks. `hatchR` can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year. In these cases, we suggest users consider how much of that data type is reasonable with their data.

For users choosing to implement `hatchR` in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record. For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.

@@

`hatchR` is built assuming data will be analyzed as daily average temperatures. Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to user's attention.

@@

We thank Laura Koller for her help designing the `hatchR` logo. Dan Isaak provided useful discussion about model development and temperature datasets.

>>>>

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**. Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for **hatchR**. The first is a R package distributed via CRAN that allows users the most customizable application of these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on **hatchR**'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface. The Shiny form trades-off some of the automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualization from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice. Finally, we provide brief summaries of general applications of **hatchR** below, but encourage users to visit articles hosted on the software webpage that extensively outline primary functions and applications, especially automating the application of predicting phenology across multiple variables and making more complex plots.

@@

The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

Fundamentally, **hatchR** assumes you have input data with two columns: a date column, giving the date (and often time) of a temperature measurement, and a temperature column, giving the associated temperature measurement (in centigrade). Other columns are okay to include, but these two columns (with any column name—just *without* spaces) are required. We expect your data to look something like this:

@@

**hatchR** assumes you've checked for missing records or errors in your data as it *will function with gaps*, so it's important to go through the data checks discussed below, as well as your own validity checks. **hatchR** can use values down to freezing (e.g, 0 °C), which returns extremely small effective values, and time to hatch or emerge may be \> 1 year. In these cases, we suggest users consider how much of that data type is reasonable with their data.

For users choosing to implement **hatchR** in program R, data can be imported from any format the user chooses, as long as users can eventually coerce their data into a `dataframe` or `tibble` format, in which each row represents a single record. For the Shiny application, users must have their data stored as a .csv (comma separated values) file for upload, which can easily be exported using datasheet software like Microsoft Excel or Google Sheets.

@@

**hatchR** is built assuming data will be analyzed as daily average temperatures. Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.

@@

We thank Laura Koller for her help designing the **hatchR** logo. Dan Isaak provided useful discussion about model development and temperature datasets.
